{\bf Practical Concurrency Testing} \\ \normalsize \vspace{1em} \begin{tabular}{c} {\em or: How I Learned to Stop Worrying and Love the Exponential Explosion} \end{tabular}

### 2018-12-08 thesis/tm.tex
tweak a bunch of smalle things in tm; confirm/adjust figure placement

\subsection{Modeling Transaction Failure}
\label{sec:tm-design-failure}

@@

\subsection{Failure Injection}
\label{sec:txn-failure}

@@

Up to four abort codes are considered depending on the testing options listed in the previous section.
All abort codes are simultaneously added to both lists,
unless already present on {\tt xabort\_codes\_ever},
in which case not added to {\tt xabort\_codes\_todo} to avoid duplicate work.
\begin{itemize}
	\item {\tt \_XABORT\_RETRY}:
		When a {\tt xbegin} preemption point is created ({\tt save\_setjmp()}),
		if {\tt -S} is not set,
		both {\tt xabort\_codes} lists are initialized to contain this code.
		This represents the possibility for a hardware transaction to fail
		for reasons outside of the programmer's control, such as system interrupts or cache eviction.
		If {\tt -S} is set they are initialized to empty,
		representing either STM's policy of failing only when a true memory conflict arises,
		or a TSX user wrapping all her {\tt \_xbegin()}s in a retry loop such as in \cite{sigbovik-htm}.
	\item {\tt \_XABORT\_CONFLICT}:
		Whenever DPOR detects a memory conflict between two transitions ({\tt shimsham\_shm()}),
		if {\tt -A} is set,
		% for each of the two which are within a transaction, % no!!!!!!!
		%it finds the nearest ancestor {\tt xbegin} preemption point by the same thread
		if the later-executed transition is a transactional success path,
		it adds this code to both lists of the immediately preceding preemption point.
		Note that a transaction should suffer a conflict abort
		only when executed {\em after} a conflicting memory access
		to avoid circular causality
		(see {\tt 410user/progs/htm\_causality.c} for rationale).
		Should the transaction happen to be executed first,
		DPOR will first try to reorder it as normal,
		and then abort it as described if the conflict persists.
	\item {\tt \_XABORT\_EXPLICIT}:
		Whenever the program invokes {\tt \_xabort()} ({\tt sched\_update\_user\_\allowbreak{}state\_machine()}),
		if {\tt -A} is set,
		this code is added to both lists, bitwise-ored with the user-supplied argument code
		as specified in \cite{htm-gcc},
		and execution of the transactional path immediately stops.
	\item {\tt \_XABORT\_CAPACITY}:
		Whenever the program invokes a system call during a transaction,
		if {\tt -A} is set,
		this code is added to both lists,
		and execution of the transactional path immediately stops.
		\revision{So called because doing so would trigger a mode switch,
		which the specification allows to abort for any reason.%
		\footnote{As far as I know, the reason for syscall aborts is both hardware- and kernel-dependent,
		and in fact nondeterministic;
		on my Haswell Core i7 running Linux 4.19, I observed a majority of {\tt \_XABORT\_CONFLICT}s,
		a scant few {\tt \_XABORT\_CAPACITY}s,
		and surprisingly many with no reason specified whatever;
		however, not once did a transaction succeed with even a single {\tt gettid()} call
		\cite{tsx-syscall}.
		For simplicity of both implementation and client programs,
		Landslide always distinguishes system call aborts with {\tt \_XABORT\_CAPACITY}.}
		See below for discussion of other cache capacity concerns.}
\end{itemize}

@@

\subsection{Data Race Analysis}
\label{sec:txn-datarace}

@@

The notation {\tt testname}$(K,N)$ will denote a test configuration of $K$ threads, each running $N$ iterations of the test logic.
All tests were run on an 8-core 2.7GHz Core i7 with 32 GB RAM.
Reported CPU-times include time spent on all state spaces
Quicksand saw fit to run,
not just the maximal or the buggy state space;
for verification tests (run with {\tt -M}),
this still includes abandoned smaller jobs that were run to saturate the set of data-race preemption points
(\Cref{chap:quicksand}).
To minimize variance in CPU-time measurements,
I ensured the test machine was not loaded beyond normal web browser use,
and ran only one instance of Landslide at a time;
for further discussion of variance see \cite{sigbovik-htm}.
The number of interleavings in each state space \revisionminor{is}, of course, deterministic and do not vary across runs.

@@

In addition to the bugs I intentionally wrote in {\tt htm1} and {\tt swapbug},
%to my pleasant surprise % only the former of these is a surprise now
Landslide also found two bugs in the ``real-world'' transactional algorithms I tested.
\begin{itemize}
	\item
	{\bf Atomicity violation.}
	{\tt avl\_insert} with any parameters higher than (2,1)
	exposed a previously-unknown bug in the transactional AVL tree.
	\Cref{fig:avlbug} shows the root cause, essentially the {\tt htm1} bug in disguise.
	This manifested
	alternately
	% landslide-trace-1533850116.527842.html
	% landslide-trace-1533850236.873032.html
	as a segfault (for test parameters (2,2) and (3,1))
	and
	% landslide-trace-1533854371.201589.html
	as a consistency-check assertion failure (for test parameters (2,3)).
	The presence of {\tt while (\_retry);} makes the necessary preemption window extremely small
	(between it and {\tt \_xbegin()}),
	making the bug extremely unlikely to manifest under stress testing,
	but Landslide is blind to such matters of chance.

	As a matter of full disclosure,
	I noted that the loop does not affect the test's possible behaviours,
	only its {\em likely} ones,
	and so removed it to make the test more Landslide-friendly.
	%keep the state space size manageable.
	%Lest any reader think I hacked the test after already knowing about the bug,
	To dispel any doubt about bias or test hacking,
	I confirmed that Landslide still finds the bug with the spin loop unmodified,
	on (3,1) in the same 53 interleavings,
	although it suffers resource exhaustion
	% from dealing with the loop
	on parameters of (2,2) or greater.
	% specifically, it runs out of file descriptors, since on bochs you need 1 fd per PP,
	% and the branch just gets 1000+ PPs deep
	\item
	{\bf Spurious spinlock abort.}
	{\tt lock\_fast} discovered a spurious transactional-path write conflict
	in the {\tt spinlock} HTM-lock implementation.%
	\footnote{{\tt lock\_fast}'s unusual (9,9) parameter shows that this state space size is constant: %in $(K,N)$:
	DPOR will always either deem all thread transitions independent and end exploration immediately,
	or the test's assertion will trip as soon as the first conflict is found.}
	This ``performance bug'' causes the lock to suffer slow-path spin-locking
	even in cases where the user's thread transitions are completely independent
	(for example, locking the root of an AVL tree,
	then traversing in different directions to make disjoint modifications),
	which the test case detects with {\tt \_xtest()}.
	\Cref{fig:spinlockbug} shows the root cause:
	the {\tt isfree()} routine (corresponding to the AVL's {\tt \_retry})
	used an atomic compare-and-swap that would always write to memory even without modifying it.
	I corrected this in the {\tt spin\_fixed} implementation by replacing it
	with a normal read
	(being used only in the transactional path, no barriers are required to protect it).
	A cursory search on Github found one user of this code,
	a transactional LevelDB implementation \cite{htm-leveldb-github},
	whose author had also noticed and corrected this problem in the same way.

@@

\begin{figure}[h]
	\begin{center}
		\begin{tabular}{l}
		\texttt{\ctype{bool} \call{hle\_spinlock\_isfree}(\ctype{spinlock\_t} *lock) \{} \\
		\texttt{~~~~\ccomment{// XXX: should be "return lock->v == 0;"}} \\
		\texttt{~~~~\flow{return} \call{\_\_sync\_bool\_compare\_and\_swap}(\&lock->v, \const{0}, \const{0});} \\
		\texttt{\}} \\
		\texttt{\ctype{void} \call{rtm\_spinlock\_acquire}(\ctype{spinlock\_t} *lock) \{} \\
		\texttt{~~~~\flow{if} ((tm\_status = \call{\_xbegin}()) == \const{\_XBEGIN\_STARTED}) \{} \\
		\texttt{~~~~~~~~\flow{if} (\call{hle\_spinlock\_isfree}(lock)) \flow{return};} \\
		\texttt{~~~~~~~~\call{\_xabort}(\const{0xff});} \\
		\texttt{~~~~\} \flow{else} \{} \\
		\texttt{~~~~~~~~\ccomment{// ... retrying \&c abbreviated for brevity ...}} \\
		\texttt{~~~~~~~~\call{hle\_spinlock\_acquire}(lock);} \\
		\texttt{~~~~\}} \\
		\texttt{\}} \\
		\texttt{\ctype{void} \call{rtm\_spinlock\_release}(\ctype{spinlock\_t} *lock)\{} \\
		\texttt{~~~~\flow{if} (\call{hle\_spinlock\_isfree}(lock)) \{} \\
		\texttt{~~~~~~~~\call{\_xend}();} \\
		\texttt{~~~~\} \flow{else} \{} \\
		\texttt{~~~~~~~~\call{hle\_spinlock\_release}(lock);} \\
		\texttt{~~~~\}} \\
		\texttt{\}} \\
		\end{tabular}
	\end{center}
	\caption[Code from {\tt spinlock-rtm.c} %, modified only to remove unrelated logic for brevity,
		showing the performance bug Landslide found.]
		{Code from {\tt spinlock-rtm.c}, modified only to remove unrelated logic for brevity,
		showing the performance bug Landslide found in {\tt lock\_fast}({\tt spinlock}).
		The {\tt isfree()} routine uses an atomic read-and-write operation where just a read would suffice,
		which leads to superfluous memory conflicts in the transactional path
		(seen at both of its callsites below).
	}
	\label{fig:spinlockbug}
\end{figure}

@@

\begin{table}[h]
	\begin{center}
		\small
		\begin{tabular}{c||c|c|c||c}
			{\bf test} & {\bf \#txn} & {\bf \#sync} & {\bf \#race} & {\bf max events verified} \\
			\hline
			{\tt htm2}		& 1	& 2	& 4	& 42	\\ % 2,3
			{\tt counter}		& 1	& 0	& 1	& 16	\\ % 2,4
			{\tt swap}		& 1	& 4	& 8	& 52	\\ % 2,2
			{\tt fig63}		& 1	& 0	& $K$-1	& 18	\\ % 3,2 k=3 -> 18; 2,4 -> 16
			{\tt avl\_insert}	& 1	& 2	& 7	& 20	\\ % 2,1 nb. only 5 drpps under -A -S but w/e
			% nb less than 9 on stm
			{\tt avl\_fixed}	& 1	& 2	& 9	& 24	\\ % 2,1 nb. again, only 5 under -A -S
			% nb stm takes it down from 13 to 10 races
			{\tt map\_basic}	& 1	& 4	& 13	& 36	\\ % 2,1
			% mysteriously had this at 6 at one point but ive no idea where that came from
			{\tt map\_basicer}	& 1	& 2	& 5	& 32	\\ % 2,2

>>>>

\subsection{Modeling transaction failure}
\label{sec:tm-design-failure}

@@

\subsection{Failure injection}
\label{sec:txn-failure}

@@

Up to four abort codes are considered depending on \cref{sec:tm-interface}'s testing options.
All abort codes are simultaneously added to both lists,
unless already present on {\tt xabort\_codes\_ever},
in which case not added to {\tt xabort\_codes\_todo} to avoid duplicate work.
\begin{itemize}
	\item {\tt \_XABORT\_RETRY}:
		When a {\tt xbegin} preemption point is created ({\tt save\_setjmp()}),
		if {\tt -S} is not set,
		both {\tt xabort\_codes} lists are initialized to contain this code.
		This represents the possibility for a hardware transaction to fail
		for reasons outside of the programmer's control, such as system interrupts or cache eviction.
		If {\tt -S} is set they are initialized to empty,
		representing either STM's policy of failing only when a true memory conflict arises,
		or a TSX user wrapping all her {\tt \_xbegin()}s in a retry loop such as in \cite{sigbovik-htm}.
	\item {\tt \_XABORT\_CONFLICT}:
		Whenever DPOR detects a memory conflict between two transitions ({\tt shimsham\_shm()}),
		if {\tt -A} is set,
		% for each of the two which are within a transaction, % no!!!!!!!
		%it finds the nearest ancestor {\tt xbegin} preemption point by the same thread
		if the later-executed transition is a transactional success path,
		it adds this code to both lists of the immediately preceding preemption point.
		Note that a transaction should suffer a conflict abort
		only when executed {\em after} a conflicting memory access
		to avoid circular causality
		(see {\tt 410user/progs/htm\_causality.c} for rationale).
		Should the transaction happen to be executed first,
		DPOR will first try to reorder it as normal,
		and then abort it as described if the conflict persists.
	\item {\tt \_XABORT\_EXPLICIT}:
		Whenever the program invokes {\tt \_xabort()} ({\tt sched\_update\_user\_\allowbreak{}state\_machine()}),
		if {\tt -A} is set,
		this code is added to both lists, bitwise-ored with the user-supplied argument code
		as specified in \cite{htm-gcc},
		and execution of the transactional path immediately stops.
	\item {\tt \_XABORT\_CAPACITY}:
		Whenever the program invokes a system call during a transaction,
		if {\tt -A} is set,
		this code is added to both lists,
		and execution of the transactional path immediately stops.
		\revision{So called because doing so would trigger a mode switch,
		which the specification allows to abort for any reason.%
		\footnote{\revision{As far as I know, the reason for syscall aborts is both hardware- and kernel-dependent,
		and in fact nondeterministic;
		on my Haswell Core i7 running Linux 4.19, I observed a majority of {\tt \_XABORT\_CONFLICT}s,
		a scant few {\tt \_XABORT\_CAPACITY}s,
		and surprisingly many with no reason specified whatever;
		however, not once did a transaction succeed with even a single {\tt gettid()} call
		\cite{tsx-syscall}.
		For simplicity of both implementation and client programs,
		Landslide always distinguishes system call aborts with {\tt \_XABORT\_CAPACITY}.}}
		See below for discussion of other cache capacity concerns.}
\end{itemize}

@@

\subsection{Data race analysis}
\label{sec:txn-datarace}

@@

The notation {\tt testname}$(K,N)$ %will denote
denotes a test configuration of $K$ threads, each running $N$ iterations of the test logic.
All tests were run on an 8-core 2.7GHz Core i7 with 32 GB RAM.
Reported CPU-times include time spent on all state spaces
Quicksand saw fit to run,
not just the maximal or the buggy state space;
for verification tests (run with {\tt -M}),
this still includes abandoned smaller jobs that were run to saturate the set of data-race preemption points
(\Cref{chap:quicksand}).
To minimize variance in CPU-time measurements,
I ensured the test machine was not loaded beyond normal web browser use,
and ran only one instance of Landslide at a time;
for further discussion of variance see \cite{sigbovik-htm}.
The number of interleavings in each state space \revisionminor{is}, of course, deterministic and do not vary across runs.

@@

In addition to the bugs I intentionally wrote in {\tt htm1} and {\tt swapbug},
%to my pleasant surprise % only the former of these is a surprise now
Landslide also found two bugs in the ``real-world'' transactional algorithms I tested.

\begin{itemize}
	\item
	{\bf Atomicity violation.}
	{\tt avl\_insert} with any parameters higher than (2,1)
	exposed a previously-unknown bug in the transactional AVL tree.
	\Cref{fig:avlbug} shows the root cause, essentially the {\tt htm1} bug in disguise.
	This manifested
	alternately
	% landslide-trace-1533850116.527842.html
	% landslide-trace-1533850236.873032.html
	as a segfault (for test parameters (2,2) and (3,1))
	and
	% landslide-trace-1533854371.201589.html
	as a consistency-check assertion failure (for test parameters (2,3)).
	The presence of {\tt while (\_retry);} makes the necessary preemption window extremely small
	(between it and {\tt \_xbegin()}),
	making the bug extremely unlikely to manifest under stress testing,
	but Landslide is blind to such matters of chance.

	As a matter of full disclosure,
	I noted that the loop does not affect the test's possible behaviours,
	only its {\em likely} ones,
	and so removed it to make the test more Landslide-friendly.
	%keep the state space size manageable.
	%Lest any reader think I hacked the test after already knowing about the bug,
	To dispel any doubt about bias or test hacking,
	I confirmed that Landslide still finds the bug with the spin loop unmodified,
	on (3,1) in the same 53 interleavings,
	although it suffers resource exhaustion
	% from dealing with the loop
	on parameters of (2,2) or greater.
	% specifically, it runs out of file descriptors, since on bochs you need 1 fd per PP,
	% and the branch just gets 1000+ PPs deep
	\item
	{\bf Spurious spinlock abort.}
	{\tt lock\_fast} discovered a spurious transactional-path write conflict
	in the {\tt spinlock} HTM-lock implementation.%
	\footnote{{\tt lock\_fast}'s unusual (9,9) parameter shows that this state space size is constant: %in $(K,N)$:
	DPOR will always either deem all thread transitions independent and end exploration immediately,
	or the test's assertion will trip as soon as the first conflict is found.}
	This ``performance bug'' causes the lock to suffer slow-path spin-locking
	even in cases where the user's thread transitions are completely independent
	(for example, locking the root of an AVL tree,
	then traversing in different directions to make disjoint modifications).
	The test case detects this with {\tt \_xtest()}.
	\Cref{fig:spinlockbug} shows the root cause:
	the {\tt isfree()} routine (corresponding to the AVL's {\tt \_retry})
	used an atomic compare-and-swap that would always write to memory even without modifying it.
	I corrected this in
	% the
	{\tt spin\_fixed}
	%implementation
	by replacing it
	with a normal read
	(being used only in the transactional path, no barriers are required to protect it).
	A cursory search on Github found one user of this code,
	a transactional LevelDB implementation \cite{htm-leveldb-github},
	whose author had also noticed and corrected this problem in the same way.

@@

\begin{figure}[t]
	\begin{center}
		\begin{tabular}{l}
		\texttt{\ctype{bool} \call{hle\_spinlock\_isfree}(\ctype{spinlock\_t} *lock) \{} \\
		\texttt{~~~~\ccomment{// XXX: should be "return lock->v == 0;"}} \\
		\texttt{~~~~\flow{return} \call{\_\_sync\_bool\_compare\_and\_swap}(\&lock->v, \const{0}, \const{0});} \\
		\texttt{\}} \\
		\texttt{\ctype{void} \call{rtm\_spinlock\_acquire}(\ctype{spinlock\_t} *lock) \{} \\
		\texttt{~~~~\flow{if} ((tm\_status = \call{\_xbegin}()) == \const{\_XBEGIN\_STARTED}) \{} \\
		\texttt{~~~~~~~~\flow{if} (\call{hle\_spinlock\_isfree}(lock)) \flow{return};} \\
		\texttt{~~~~~~~~\call{\_xabort}(\const{0xff});} \\
		\texttt{~~~~\} \flow{else} \{} \\
		\texttt{~~~~~~~~\ccomment{// ... retrying \&c abbreviated for brevity ...}} \\
		\texttt{~~~~~~~~\call{hle\_spinlock\_acquire}(lock);} \\
		\texttt{~~~~\}} \\
		\texttt{\}} \\
		\texttt{\ctype{void} \call{rtm\_spinlock\_release}(\ctype{spinlock\_t} *lock)\{} \\
		\texttt{~~~~\flow{if} (\call{hle\_spinlock\_isfree}(lock)) \{} \\
		\texttt{~~~~~~~~\call{\_xend}();} \\
		\texttt{~~~~\} \flow{else} \{} \\
		\texttt{~~~~~~~~\call{hle\_spinlock\_release}(lock);} \\
		\texttt{~~~~\}} \\
		\texttt{\}} \\
		\end{tabular}
	\end{center}
	\caption[Code from {\tt spinlock-rtm.c} %, modified only to remove unrelated logic for brevity,
		showing the performance bug Landslide found.]
		{Code from {\tt spinlock-rtm.c}, modified only to remove unrelated logic for brevity,
		showing the performance bug Landslide found in {\tt lock\_fast}({\tt spinlock}).
		The {\tt isfree()} routine uses an atomic read-and-write operation where just a read would suffice,
		which leads to superfluous memory conflicts in the transactional path
		(seen at both of its callsites below).
	}
	\label{fig:spinlockbug}
\end{figure}

@@

% TODO: check figure placement
\begin{table}[t]
	\begin{center}
		\small
		\begin{tabular}{c||c|c|c||c}
			{\bf test} & {\bf \#txn} & {\bf \#sync} & {\bf \#race} & {\bf max events verified} \\
			\hline
			{\tt htm2}		& 1	& 2	& 4	& 42	\\ % 2,3
			{\tt counter}		& 1	& 0	& 1	& 16	\\ % 2,4
			{\tt swap}		& 1	& 4	& 8	& 52	\\ % 2,2
			{\tt fig63}		& 1	& 0	& $K$-1	& 18	\\ % 3,2 k=3 -> 18; 2,4 -> 16
			{\tt avl\_insert}	& 1	& 2	& 7	& 20	\\ % 2,1 nb. only 5 drpps under -A -S but w/e
			% nb less than 9 on stm
			{\tt avl\_fixed}	& 1	& 2	& 9	& 24	\\ % 2,1 nb. again, only 5 under -A -S
			% nb stm takes it down from 13 to 10 races
			{\tt map\_basic}	& 1	& 4	& 13	& 36	\\ % 2,1
			% mysteriously had this at 6 at one point but ive no idea where that came from
			{\tt map\_basicer}	& 1	& 2	& 5	& 32	\\ % 2,2
